DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life

### 2022-03-08 manuscript/manuscript.tex
beefed methods

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother
\shorttitle{DateLife: revealing the dated Tree of Life}
\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 3542}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

@@

Chronograms --phylogenies with branch lengths proportional to time, provide key data for the study of natural processes in many areas of biological research, such as developmental biology (cite) to conservation biology (Felsenstein, 1985; Campbell O. Webb, 2000), historical biogeography, a species diversification (Morlon, 2014; Posadas, Crisci, \& Katinas, 2006).

@@

\hypertarget{the-datelife-algorithm}{%
\section{The DateLife algorithm}\label{the-datelife-algorithm}}

\emph{\textbf{Providing an input.--}} DateLife starts with an input query consisting of at least two taxon names, which can be provided as a comma separated character string, or as tip labels on a tree. If the input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a ``phylo'' R object (Paradis, Claude, \& Strimmer, 2004). The input tree is not required to have branch lengths, and its topology is used on the summary steps described below.

Currently, DateLife only accepts scientific names as input, and they can belong to any inclusive taxonomic group (e.g., genus, family, tribe, etc.) or binomial specific. Subspecies and variants are ignored. If an input taxon name belongs to an ``inclusive'' taxonomic group the algorithm has two alternative behaviors defined by the ``get species from taxon'' flag. If the flag is active, the DateLife algorithm retrieves all species names within the ``inclusive'' taxonomic group and adds them to the input. If the flag is inactive, DateLife ignores the ``inclusive'' taxon names from the input.

Input scientific names are processed using the Taxonomic Name Resolution Service (TNRS; Boyle et al. (2013)) provided by OpenTree (OpenTreeOfLife et al., n.d.). TNRS increases the probability of correctly finding the query organisms in the chronogram database, as it detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard.

The processed input taxon names are saved as a special R object (of a newly defined class \texttt{datelifeQuery}) that contains the processed names, the corresponding taxonomic id numbers, and the topology of the input tree if any was provided. The \texttt{datelifeQuery} object is used next to search the chronogram database.

\emph{\textbf{Searching the database.--}} Chronograms in the DateLife database with at least two matching processed taxon names are identified and pruned down to preserve only the matched names as tips. Then, each pruned chronogram is transformed to a patristic distance matrix. This format facilitates and greatly speeds up all downstream analyses and summaries. The matrices are associated to the citation of the original study and stored as an R object of class \texttt{datelifeResult}.

\emph{\textbf{Summarizing search results.--}} At this point, summary data is collected to inform decisions for ulterior steps in the user workflow. Types of summary information provided are:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The pruned source chronograms are used to get pairwise node ages of your query taxa (Figure \ref{fig:workflow} subfigure X).
\item
  Age of the root of pruned source chronograms. This can correspond to the mrca age of your group if the pruned source chronogram has all your query taxa. If not, the root correponds to the mrca of subgroup withing your query taxa.
\item
  Citations of studies where pruned source chronograms were originally published.
\item
  A summary table with all of the above.
\item
  A single chronogram summarizing all pruned source chronograms, or a subset defined by the user.
  To construct summary trees, \texttt{datelife} requires a fixed topology. Users can provide one from the literature, or one of their own making. If no topology is provided, \texttt{datelife} chooses one automatically retrieved from an expert phylogenetic information database such as the OpenTree synthetic tree.
\end{enumerate}

????? More meat on the methods needed: e.g.~how the distances are generated and summarized, how conflict is handled\ldots{}

Then we summarize source chronograms into a single patristic distance matrix using a method chosen by the user. Summarizing method options available include Super Distance Matrix method (SDM, Criscuolo, Berry, Douzery, \& Gascuel, 2006) and any summary statistics such as median, minimum and maximum ages. Finally, \texttt{datelife} applies the summarized time distances as node calibrations to date the given topology with BLADJ (Campbell O. Webb \& Donoghue, 2005).

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{5}
\tightlist
\item
  A report of successful matches of input taxon names across pruned source chronograms.
\item
  A single pruned source chronogram with the most matching input taxon names.
\end{enumerate}

If there is no information available for any input taxon name, users can also create both age and phylogenetic data for the missing branches with a variety of algorithms, ??? like how?

\emph{\textbf{Visualizing summaries.--}} Users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R ``phylo'' format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s graphic generation functions.

The DateLife algorithm illustrated in figure \ref{fig:workflow} is fully implemented using the R language. The latest stable version of the R package \texttt{datelife} is available from the CRAN repository (v0.6.2; Sanchez-Reyes et al. (2022)), and relies on functionalities from various biological R packages:
ape (Paradis et al., 2004),
bold (Chamberlain et al., 2019),
geiger (Harmon, Weir, Brock, Glor, \& Challenger, 2008),
paleotree (Bapst, 2012),
phyloch (Heibl, 2008),
phylocomr (Ooms \& Chamberlain, 2018),
phytools (Revell, 2012),
rotl (Michonneau, Brown, \& Winter, 2016), and
taxize (Chamberlain \& Szöcs, 2013; Chamberlain et al., 2019).

@@

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{case-study}{%
\subsection{Case study}\label{case-study}}

@@

\hypertarget{cross-validation-test}{%
\subsection{Cross-validation test}\label{cross-validation-test}}

Data from source chronograms can be also used to date tree topologies with no branch lengths, as well as trees with branch lengths as relative substitution rates (Figs. \ref{fig:cvbladj} and \ref{fig:cvbold}). As a form of cross validation, we took tree topologies from each study and calibrated them using time of lineage divergence data from all other source chronograms. In the absence of branch lengths, the ages of internal nodes were recovered with a high precision in almost all cases (except for studies 3, and 5; Fig. \ref{fig:cvbladj}).
Maximum tree ages were only recovered in one case (study 2; Fig. \ref{fig:cvbladj}).
We also demonstrate the usage of PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007) as an alternative method to BLADJ. For this, we run a \texttt{datelife} branch length reconstruction that searches for DNA sequence data from the Barcode of Life Data System {[}BOLD; Ratnasingham and Hebert (2007){]} to generate branch lengths. We were able to successfully generate a tree with BOLD branch lengths for all of the Fringillidae source chronograms. However, dating with PATHd8 using congruified calibrations, was only successful in
three cases (studies 3, 5, and 9, shown in Fig. \ref{fig:cvbold}). From these, two trees have a different sampling than the original source chronogram, mainly because DNA BOLD data for some species is absent from the database. Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
More examples and code used to generate these trees were developed on an open repository that is available for consultation and reuse at \url{https://github.com/LunaSare/datelife_examples}.

@@

At the time of writing of this manuscript (Mar 07, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015). A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time. As chronograms are added to Phylesystem, they are incorporated into an updated \texttt{datelife}'s database that is assigned a new version number, followed by a package release on CRAN. \texttt{datelife}'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
Users can also upload new chronograms to OpenTree themselves, and trigger an update of the \texttt{datelife} database to incorporate the new chronograms, to have them immediately available for analysis.

@@

In other areas of biological research, such as ecology and conservation biology, it has been shown that at least some data on lineage divergence represents a relevant improvement for testing alternative hypothesis using phylogenetic distance (Campbell O. Webb, Ackerly, \& Kembel, 2008). Hence, we integrated into datelife's workflow different ways of creating branch lengths in the absence of starting branch length information for taxa lacking this information (using the BLADJ option). ``Making up'' branch lengths is an accepted practice in scientific publications: Jetz, Thomas, Joy, Hartmann, and Mooers (2012), created a time-calibrated tree of all 9,993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al. (2018) created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37\% had molecular data; Smith and Brown (2018) constructed a tree of 353,185 seed plants where only 23\% had molecular data. Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration with the polytomy resolution and branch assignation methods. There has yet to be a thorough analysis of what can go wrong when one goes beyond the data in this way, so we urge caution; we also urge readers to follow the example of many of the large tree papers cited above and make sure results are substantially similar between trees fully reconstructed with molecular or other data, and trees that are reconstructed using taxonomy by resolving polytomies at random following a statistical model.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-antonelli2017supersmart}{}}%
Antonelli, A., Hettling, H., Condamine, F. L., Vos, K., Nilsson, R. H., Sanderson, M. J., \ldots{} Vos, R. A. (2017). {Toward a self-updating platform for estimating rates of speciation and migration, ages, and relationships of Taxa}. \emph{Systematic Biology}, \emph{66}(2), 153--166. \url{https://doi.org/10.1093/sysbio/syw066}

@@

@@

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-opentreeAPIs}{}}%
OpenTreeOfLife, Redelings, B., Cranston, K. A., Allman, J., Holder, M. T., \& McTavish, E. J. (n.d.). {Open Tree of Life APIs v. 3.0}. Retrieved from \url{\%7Bhttps://github.com/OpenTreeOfLife/germinator/wiki/Open-Tree-of-Life-Web-APIs\%7D}

@@

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-smith2018constructing}{}}%
Smith, S. A., \& Brown, J. W. (2018). Constructing a broadly inclusive seed plant phylogeny. \emph{American Journal of Botany}, \emph{105}(3), 302--314.

@@

\caption{Stylized DateLife workflow. This shows the general workflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website  at \url{www.datelife.org/query/}. Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.}

>>>>

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother
\shorttitle{DateLife: revealing the dated Tree of Life}
\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 3863}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

@@

Chronograms --phylogenies with branch lengths proportional to time-- provide key data for the study of natural processes in many areas of biological research, such as developmental biology (Delsuc et al., 2018; Laubichler \& Maienschein, 2009), conservation biology (Felsenstein, 1985; Campbell O. Webb, 2000), historical biogeography (Posadas, Crisci, \& Katinas, 2006), and species diversification (Magallon \& Sanderson, 2001; Morlon, 2014).

@@

\hypertarget{description}{%
\section{Description}\label{description}}

The DateLife algorithm is fully implemented using the R language. The latest stable version of the R package \texttt{datelife} is available from the CRAN repository (v0.6.2; Sanchez-Reyes et al. (2022)), and relies on functionalities from various biological R packages:
ape (Paradis, Claude, \& Strimmer, 2004),
bold (Chamberlain et al., 2019),
geiger (Harmon, Weir, Brock, Glor, \& Challenger, 2008),
paleotree (Bapst, 2012),
phyloch (Heibl, 2008),
phylocomr (Ooms \& Chamberlain, 2018),
phytools (Revell, 2012),
rotl (Michonneau, Brown, \& Winter, 2016), and
taxize (Chamberlain \& Szöcs, 2013; Chamberlain et al., 2019).
Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife algorithm: providing an input, searching a chronogram database, and summarizing results from the search.

\hypertarget{providing-an-input}{%
\subsection{Providing an input}\label{providing-an-input}}

DateLife starts with an input query consisting of at least two taxon names, which can be provided as a comma separated character string, or as tip labels on a tree. If the input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a ``phylo'' R object (Paradis et al., 2004). The input tree is not required to have branch lengths, and its topology is used on the summary steps described below.

Currently, DateLife only accepts scientific names as input, and they can belong to any inclusive taxonomic group (e.g., genus, family, tribe, etc.) or binomial specific. Subspecies and variants are ignored. If an input taxon name belongs to an inclusive taxonomic group the algorithm has two alternative behaviors defined by the ``get species from taxon'' flag. If the flag is active, the DateLife algorithm retrieves all species names within the inclusive taxonomic group and adds them to the input. If the flag is inactive, DateLife ignores the inclusive taxon names from the input.

Input scientific names are processed using the Taxonomic Name Resolution Service (TNRS; Boyle et al. (2013)), which increases the probability of correctly finding the queried taxon names in the chronogram database. TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard. DateLife implements TNRS using OpenTree's taxonomic as standard (Open Tree Of Life et al., 2016; Rees \& Cranston, 2017).

The processed input taxon names are saved as an R object of a newly defined class \texttt{datelifeQuery} that is used in the following steps. It contains the processed names, the corresponding OpenTree taxonomic id numbers, and the topology of the input tree if any was provided.

\hypertarget{searching-the-database}{%
\subsection{Searching the database}\label{searching-the-database}}

A DateLife search consists of matching processed taxon names to tip labels in a chronogram database. Chronograms with at least two matching tip labels are identified and pruned down to preserve only the matched tips.

Matching pruned chronograms are stored as individual patristic distance matrices (Figure \ref{fig:workflow} subfigure X). This format speeds up extraction of pairwise taxon ages of the queried taxa, as opposed to searching the ancestor node of a pair of taxa in a ``phylo'' object or newick string.
The patristic matrices are also associated to the study citation where the original chronogram was published, and stored as an R object of the newly defined class \texttt{datelifeResult}.

DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies. It is constructed from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert phylogenetic knowledge with rich metadata (McTavish et al., 2015) that allows automatic and reproducible construction of a chronogram database.

\hypertarget{summarizing-search-results}{%
\subsection{Summarizing search results}\label{summarizing-search-results}}

At this point, summary information is extracted from the \texttt{datelifeResult} object to inform decisions for ulterior steps in the user workflow. Age data from matching pruned chronograms is summarized and used to generate a single summary chronogram. Other basic summary information available is:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The matching pruned chronograms as newick strings or ``phylo'' objects.
\item
  The ages of the root of all matching pruned chronograms. This can correspond to the age of the most recent common ancestor (mrca) of your group of interest if the pruned chronograms have all taxa belonging to the group. If not, the root corresponds to the mrca ofa subgroup withing your group of interest.
\item
  Study citations where original chronograms were published.
\item
  A report of input taxon names matches across pruned chronograms.
\item
  The single matching pruned chronogram with the most input taxon names.
\end{enumerate}

\emph{\textbf{Identifying groves.--}} To generate a single summary chronogram, the DateLife algorithm starts by identifying the matching pruned chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al. (2009). In rare cases, a group of trees can have multiple groves. By default, DateLife chooses the grove with the most taxa, however, the ``criterion = trees'' flag allows the user to choose the grove with the most trees instead.

\emph{\textbf{Choosing a topology.--}} Due to sampling, DateLife requires a tree topology to summarize age data upon it. Users can provide one as input from the literature, or one of their own making. If no topology is provided, DateLife automatically subsets one from the OpenTree synthetic tree (Open Tree Of Life et al., 2019).

DateLife can also reconstruct branch lengths proportional to substitution rates on the fixed tree topology from available genetic data from BOLD.

\emph{\textbf{Congruifying nodes.--}} DateLife then implements the congruification method (Eastman, Harmon, \& Tank, 2013) to find nodes belonging to the same clade across matching pruned chronograms.
Congruified node ages stored as a \texttt{congruifiedCalibrations} object are then matched to nodes in the chosen tree topology and stored as a \texttt{matchedCalibrations} object.

\emph{\textbf{Summarizing node ages.--}} DateLife summarizes matched calibrations into a single patristic distance matrix using different methods. Summarizing options implemented include Super Distance Matrix method (SDM, Criscuolo, Berry, Douzery, \& Gascuel, 2006) and summary statistics such as median, minimum and maximum ages.

\emph{\textbf{Dating the tree topology.--}} Summarized calibrations can be applied as secondary calibrations with different dating methods currently supported within DateLife: MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007), BLADJ (Campbell O. Webb, Ackerly, \& Kembel, 2008; Campbell O. Webb \& Donoghue, 2005), and treePL (Stephen A. Smith \& O'Meara, 2012).
By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm that assigns ages to nodes with no data evenly between nodes with age data, which minimizes age variance in the resulting chronogram (Campbell O. Webb et al., 2008).
When there is conflict in ages across node with age data, the algorithm ignores ages that are older than parent nodes and/or younger than descendant nodes.

If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature. If none is provided, DateLife assigns an arbitrary age to the root as 10\% older than the oldest age available within the group.

\emph{\textbf{Visualizing results.--}}
Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R ``phylo'' format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s chronogram plot generation functions.

@@

\hypertarget{case-study}{%
\section{Case study}\label{case-study}}

@@

\hypertarget{cross-validation-test}{%
\section{Cross-validation test}\label{cross-validation-test}}

Data from source chronograms can be also used to date tree topologies with no branch lengths, as well as trees with branch lengths as relative substitution rates (Figs. \ref{fig:cvbladj} and \ref{fig:cvbold}). As a form of cross validation, we took tree topologies from each study and calibrated them using time of lineage divergence data from all other source chronograms. In the absence of branch lengths, the ages of internal nodes were recovered with a high precision in almost all cases (except for studies 3, and 5; Fig. \ref{fig:cvbladj}).
Maximum tree ages were only recovered in one case (study 2; Fig. \ref{fig:cvbladj}).
We also demonstrate the usage of PATHd8 (Britton et al., 2007) as an alternative method to BLADJ. For this, we run a \texttt{datelife} branch length reconstruction that searches for DNA sequence data from the Barcode of Life Data System {[}BOLD; Ratnasingham and Hebert (2007){]} to generate branch lengths. We were able to successfully generate a tree with BOLD branch lengths for all of the Fringillidae source chronograms. However, dating with PATHd8 using congruified calibrations, was only successful in
three cases (studies 3, 5, and 9, shown in Fig. \ref{fig:cvbold}). From these, two trees have a different sampling than the original source chronogram, mainly because DNA BOLD data for some species is absent from the database. Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
More examples and code used to generate these trees were developed on an open repository that is available for consultation and reuse at \url{https://github.com/LunaSare/datelife_examples}.

@@

At the time of writing of this manuscript (Mar 08, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015). A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time. As chronograms are added to Phylesystem, they are incorporated into an updated \texttt{datelife}'s database that is assigned a new version number, followed by a package release on CRAN. \texttt{datelife}'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
Users can also upload new chronograms to OpenTree themselves, and trigger an update of the \texttt{datelife} database to incorporate the new chronograms, to have them immediately available for analysis.

@@

In other areas of biological research, such as ecology and conservation biology, it has been shown that at least some data on lineage divergence represents a relevant improvement for testing alternative hypothesis using phylogenetic distance (Campbell O. Webb et al., 2008). Hence, we integrated into datelife's workflow different ways of creating branch lengths in the absence of starting branch length information for taxa lacking this information (using the BLADJ option). ``Making up'' branch lengths is an accepted practice in scientific publications: Jetz, Thomas, Joy, Hartmann, and Mooers (2012), created a time-calibrated tree of all 9,993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al. (2018) created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37\% had molecular data; Stephen A. Smith and Brown (2018) constructed a tree of 353,185 seed plants where only 23\% had molecular data. Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration with the polytomy resolution and branch assignation methods. There has yet to be a thorough analysis of what can go wrong when one goes beyond the data in this way, so we urge caution; we also urge readers to follow the example of many of the large tree papers cited above and make sure results are substantially similar between trees fully reconstructed with molecular or other data, and trees that are reconstructed using taxonomy by resolving polytomies at random following a statistical model.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-ane2009groves}{}}%
Ané, C., Eulenstein, O., Piaggio-Talice, R., \& Sanderson, M. J. (2009). Groves of phylogenetic trees. \emph{{Annals of Combinatorics}}, \emph{13}(2), 139--167.

\leavevmode\vadjust pre{\hypertarget{ref-antonelli2017supersmart}{}}%
Antonelli, A., Hettling, H., Condamine, F. L., Vos, K., Nilsson, R. H., Sanderson, M. J., \ldots{} Vos, R. A. (2017). {Toward a self-updating platform for estimating rates of speciation and migration, ages, and relationships of Taxa}. \emph{Systematic Biology}, \emph{66}(2), 153--166. \url{https://doi.org/10.1093/sysbio/syw066}

@@

\leavevmode\vadjust pre{\hypertarget{ref-delsuc2018phylogenomic}{}}%
Delsuc, F., Philippe, H., Tsagkogeorga, G., Simion, P., Tilak, M.-K., Turon, X., \ldots{} Douzery, E. J. (2018). A phylogenomic framework and timescale for comparative studies of tunicates. \emph{BMC Biology}, \emph{16}(1), 1--14.

\leavevmode\vadjust pre{\hypertarget{ref-Eastman2013}{}}%
Eastman, J. M., Harmon, L. J., \& Tank, D. C. (2013). {Congruification: Support for time scaling large phylogenetic trees}. \emph{Methods in Ecology and Evolution}, \emph{4}(7), 688--691. \url{https://doi.org/10.1111/2041-210X.12051}

@@

\leavevmode\vadjust pre{\hypertarget{ref-Huelsenbeck2001}{}}%
Huelsenbeck, J. P., \& Ronquist, F. (2001). {MRBAYES: Bayesian inference of phylogenetic trees}. \emph{Bioinformatics}, \emph{17}(8), 754--755. \url{https://doi.org/10.1093/bioinformatics/17.8.754}

@@

\leavevmode\vadjust pre{\hypertarget{ref-laubichler2009form}{}}%
Laubichler, M. D., \& Maienschein, J. (2009). \emph{Form and function in developmental evolution}. Cambridge University Press.

\leavevmode\vadjust pre{\hypertarget{ref-magallon2001absolute}{}}%
Magallon, S., \& Sanderson, M. J. (2001). Absolute diversification rates in angiosperm clades. \emph{Evolution}, \emph{55}(9), 1762--1780.

@@

\leavevmode\vadjust pre{\hypertarget{ref-opentreeAPIs}{}}%
Open Tree Of Life, Redelings, B., Cranston, K. A., Allman, J., Holder, M. T., \& McTavish, E. J. (2016). {Open Tree of Life APIs v3.0}. \emph{{Open Tree of Life Project}}, (Online Resources). Retrieved from \url{\%7Bhttps://github.com/OpenTreeOfLife/germinator/wiki/Open-Tree-of-Life-Web-APIs\%7D}

\leavevmode\vadjust pre{\hypertarget{ref-opentreeoflife2019synth}{}}%
Open Tree Of Life, Redelings, B., Sánchez Reyes, L. L., Cranston, K. A., Allman, J., Holder, M. T., \& McTavish, E. J. (2019). Open tree of life synthetic tree v12.3. \emph{Zenodo}. Retrieved from \url{https://doi.org/10.5281/zenodo.3937742}

@@

\leavevmode\vadjust pre{\hypertarget{ref-rees2017automated}{}}%
Rees, J. A., \& Cranston, K. (2017). Automated assembly of a reference taxonomy for phylogenetic data synthesis. \emph{Biodiversity Data Journal}, (5).

@@

\leavevmode\vadjust pre{\hypertarget{ref-Ronquist2003}{}}%
Ronquist, F., \& Huelsenbeck, J. P. (2003). {MrBayes 3: Bayesian phylogenetic inference under mixed models}. \emph{Bioinformatics}, \emph{19}(12), 1572--1574. \url{https://doi.org/10.1093/bioinformatics/btg180}

@@

\leavevmode\vadjust pre{\hypertarget{ref-smith2018constructing}{}}%
Smith, Stephen A., \& Brown, J. W. (2018). Constructing a broadly inclusive seed plant phylogeny. \emph{American Journal of Botany}, \emph{105}(3), 302--314.

\leavevmode\vadjust pre{\hypertarget{ref-Smith2012}{}}%
Smith, Stephen A., \& O'Meara, B. C. (2012). {TreePL: Divergence time estimation using penalized likelihood for large phylogenies}. \emph{Bioinformatics}, \emph{28}(20), 2689--2690. \url{https://doi.org/10.1093/bioinformatics/bts492}

@@

\caption{Stylized DateLife algorithm showing the main steps that are performed with the R package \texttt{datelife}, or online at \url{www.datelife.org/query/}. Details on the functions involved on each workflow are available at \texttt{datelife}'s R package manual https://phylotastic.org/datelife/articles/.}
